When did a variant start spreading?

### 2022-07-06 notes/variant_origin.tex
multi-variant scribbles

These quantities can be calculated recursively:
\begin{equation}
    R(t-1)_n = \sum_m R(t)_m g_{m,n} \quad \mathrm{with} \quad R(T)_n = p(m_{T}|n,\epsilon)
\end{equation}
\begin{equation}
    Q(t+1,\tau)_n = \sum_m g_{n,m}Q(t,\tau)_m  \quad \mathrm{with} \quad Q(\tau,\tau)_n = \delta_{n,1}
\end{equation}

@@

\begin{equation}
    P(\{m_t\}, \{c_t\}) = \sum_{\tau_0}^{t_0}\sum_{\tau_d=\tau_0+1}^{t_d} R(\tau_d)_n (1-e^{-n\mu L}) Q(\tau_d,\tau)_n R_d(\tau_d)
\end{equation}
where $\tau_d$ is the seeding time of the daughter clone, $\tau_0$ is the seeding of the parent clone, and $\{m_t\}, \{c_t\}$ are the parent and daughter cases count trajectories.
The probability of being seeded in generation $\tau_0$ is then
\begin{equation}
    P(\tau_0) \sim \sum_{\tau_d=\tau_0+1}^{t_d} R(\tau_d)_n (1-e^{-n\mu L}) Q(\tau_d,\tau)_n R_d(\tau_d)
\end{equation}

>>>>

These quantities can be calculated recursively:
\begin{equation}
    R(t-1)_n = \sum_k R(t)_k g_{k,n} p(m_{t-1}|n,\epsilon) \quad \mathrm{with} \quad R(T)_n = p(m_{T}|n,\epsilon)
\end{equation}
\begin{equation}
    Q(t+1,\tau)_n = \sum_k g_{n,k} p(m_t|k,\epsilon) Q(t,\tau)_k  \quad \mathrm{with} \quad Q(\tau,\tau)_n = \delta_{n,1}
\end{equation}

@@

\begin{equation}
    P(\{m_t\}, \{c_t\}) = \sum_{\tau_0}^{t_0}\sum_{\tau_d=\tau_0+1}^{t_d} R(\tau_d)_n (1-e^{-n\mu L}) Q(\tau_d,\tau)_n R_d(\tau_d)
\end{equation}
where $\tau_d$ is the seeding time of the daughter clone, $\tau_0$ is the seeding of the parent clone, and $\{m_t\}, \{c_t\}$ are the parent and daughter cases count trajectories.
The probability of being seeded in generation $\tau_0$ is then
\begin{equation}
    P(\tau_0) \sim \sum_{\tau_d=\tau_0+1}^{t_d} \sum_n R(\tau_d)_n (1-e^{-n\mu L}) Q(\tau_d,\tau)_n R_d(\tau_d)
\end{equation}

\subsection*{Mutation categories}
Instead of tracking clones, we might want to track genotypes with 0,1,2,... mutations.
In this case, mutation happen independently and there is a constant flux of cases from lower to higher mutation classes.
The last "catch-all" class grows with a rate that is $\mu L$ higher than the rest.

In this case, we'd start with the most mutated class and would need to extend the propagator to include the mutational influx.
The trajectory of mutant class $k$ is conditional on the mutant class $k-1$.
\begin{equation}
    \begin{split}
        P(\{m^0_t\}, \{m^1_t\}, \{m^2_t\} \ldots | \tau) & = \prod_{t=\tau+1}^T\sum_{n^0_t=1}\sum_{n^1_t=0}^\infty\sum_{n^2_t=0}^\infty p(m^0_t|n^0_t,\epsilon)g_{n^0_t,n^0_{t-1}} p(m^1_t|n^1_t,\epsilon) g_{n^1_t,n^1_{t-1}, n^0_{t-1}}p(m^2_t|n^2_t,\epsilon)g_{n^2_t,n^2_{t-1}, n^1_{t-1}} \\
         & = \sum_{n^0_\tau=1}\sum_{n^1_t=0}^\infty\sum_{n^2_t=0}^\infty p(m^0_\tau|n^0_\tau,\epsilon) p(m^1_\tau|n^1_\tau,\epsilon) p(m^2_\tau|n^2_\tau,\epsilon) \\
          &   \ldots \\
          & \sum_{n^0_{T-1}=1}  p(m^0_{T-1}|n^0_{T-1},\epsilon) g_{n^0_{T-1},n^0_{T-2}} \sum_{n^1_{T-1}=0}^\infty p(m^1_{T-1}|n^1_{T-1},\epsilon) g_{n^1_{T-1},n^1_{T-2}, n^0_{T-2}}  \sum_{n^2_{T-1}=0}^\infty p(m^2_{T-1}|n^2_{T-1},\epsilon) g_{n^2_{T-1},n^2_{T-2}, n^1_{T-2}} \\
          &   \sum_{n^0_T=1}p(m^0_T|n^0_T,\epsilon)  g_{n^0_T,n^0_{T-1}} \sum_{n^1_T=0}^\infty p(m^1_T|n^1_T,\epsilon)g_{n^1_T,n^1_{T-1}, n^0_{T-1}}\sum_{n^2_T=0}^\infty p(m^2_T|n^2_T,\epsilon)  g_{n^2_T,n^2_{T-1}, n^1_{T-1}}
    \end{split}
\end{equation}
The quantity $g_{n^2_t,n^2_{t-1}, n^1_{t-1}}$ describes how many double mutated cases are present in generation $t$ if there where $n^2_{t-1}$ double mutated and $n^1_{t-1}$ single mutated individuals previously.
The mean number of progeny is then $\rho e^{-\mu L} (n^2_{t-1} + \mu L n^1_{t-1})$.
This problem can again be solved recursively by defining $R^k(t)_{n^k,n^{k-1}}$ of mutant category $k$ to be
\begin{equation}
    R(t-1)_{n^k,n^{k-1}} = \sum_{l^k} R(t)_{l^k,l^{k-1}} g_{k,n} p(m_{t-1}|n,\epsilon) \quad \mathrm{with} \quad R(T)_n = p(m_{T}|n,\epsilon)
\end{equation}
